---
title: "Latest update for upswings"
author: "Quang Nguyen"
date: "2022-03-03"
output:
    html_document:
        keep_md: yes
        toc: true
        toc_float: true
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setting up  
```{r, message=FALSE}
library(here)
library(covidcast)
library(epiprocess)
library(tidyverse)
library(ggsci)
library(tsibble)
library(patchwork)
here::i_am(path = "notebooks/01-latest-upswings.Rmd")
source(here("R", "utils.R"))
(settings <- get_settings(start_date = "2021-01-01", end_date = "2021-12-31"))
```
# Load data 
```{r load_data, cache=TRUE, message=TRUE}
cases <- covidcast_signal(data_source = "jhu-csse", 
                          signal = "confirmed_incidence_num", 
                          start_day = settings$start_date, 
                          end_day = settings$end_date, 
                          geo_type = "state",
                          geo_values = c("ma"),
                          as_of = Sys.Date())

(df <- as_epi_df(cases, geo_type = "state", time_type = "day", as_of = max(cases$issue)) %>% 
        select(geo_value, time_value, cases = value) %>% as_tsibble() %>%
        index_by(epiweek = ~yearweek(., week_start = 7)) %>% 
        group_by(geo_value) %>% 
        summarise(cases = sum(cases)) %>% ungroup() %>% 
        as_tibble() %>% rename(time_value = epiweek) %>% 
        as_epi_df(geo_type = "state", time_type = "day", as_of = max(cases$issue))
        
)


```

# Compute upswings on real data  

The `rel_change` method in the `growth_rate` function from `epiprocess` defines relative change at focal time $T$ with bandwith $h$ as: 

$$\frac{1}{h} * \left(\frac{\bar{B}}{\bar{A}} - 1\right) = \frac{1}{h} * \left(\frac{\bar{B} - \bar{A}}{\bar{A}}\right) = \\ \frac{1}{h} * \left(\frac{(h/2)^{-1}\left(\sum_{t = T+1}^{T + h/2} Y_t - \sum_{t = T+1-h/2}^{T} Y_t\right)}{(h/2)^{-1}\sum_{t = T+1-h/2}^{T} Y_t}\right) = \frac{1}{h} R^{h/2}_{T + h/2}$$

where $R_{T+h/2}^{h/2}$ is the **the actual $h/2$-epiweek-incidence relative change** as defined in notebook 7 on computing upswings on incidence data. 

Using the `rel_change` option in the `growth_rate` function from `epiprocess` with `h` being the set threshold   

```{r compute_upswing, eval=FALSE}
test_df <- df %>% mutate(c_gr_raw = growth_rate(time_value, cases, method = "rel_change", h = 2), 
                         c_gr = c_gr_raw * 2) %>%  
    mutate(
        prev_cumulative = lag(cases, order_by = time_value, n = 7 * h),
        h_ew_inc = cases - prev_cumulative,
        prev_h_ew_inc = lag(h_ew_inc, order_by = time_value, n = 7 * h),
        rel_change = (h_ew_inc - prev_h_ew_inc) / prev_h_ew_inc,
        surge = case_when(rel_change >= surge_thresh & h_ew_inc >= min_thresh * h ~ TRUE,
                          rel_change < surge_thresh & h_ew_inc >= min_thresh * h ~ FALSE,
                          h_ew_inc < min_thresh * h ~ NA),
        prev_surge = lag(surge, order_by = time_value, n = 7 * h),
        cumulative = cases,
        inc = cases - lag(cases, order_by = time_value, n = 7),
        target_end_date = time_value
    )

p1 <- test_df %>% ggplot(aes(x = time_value, y = c_gr_raw)) + geom_point() 

p2 <- test_df %>% ggplot(aes(x = time_value, y = c_gr)) + geom_point()


p1 + p2
```




