---
title: "Evaluating the multiple forecasters"
author: "Quang Nguyen"
date: "Last compiled on `r format(Sys.time(), '%Y-%m-%d')`"
output: 
    #github_document: default
    html_document:
        keep_md: yes
        toc: true
        toc_float: true
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data pre-loading and processing

```{r, message=FALSE, warning=FALSE}
library(here)
library(covidcast)
library(epiprocess)
library(zoltr)
library(tidyverse)
library(ggsci)
library(tsibble)
library(covidHubUtils)
library(lubridate)
library(rlang)
library(patchwork)
library(pROC)
here::i_am(path = "notebooks/multiple-forecaster.Rmd")
source(here("R", "utils.R"))
# (settings <- get_settings(start_date = "2020-06-01", end_date = "2022-03-01"))
theme_set(theme_bw())
```

First let's set some parameters and constants.  

```{r}
d_range <- seq(ymd("2020-11-01"), ymd("2022-06-01"), by = 7)
curr_date <- "2022-06-01"
h <- 4
wk_ahead <- 1
inc_case_targets <- paste(1:h, "wk ahead inc case")
surge_thresh <- 0.5
min_inc <- 20
```

# Loading forecasts and underlying data

Using `covidHubUtils` and `zoltar` we load underlying data as well as forecasts for models available in the US hub. Here we retrieve US states incident case forecasts 1-4 weeks in advance.  

```{r, cache=TRUE}
models <- c("COVIDhub-baseline", "RobertWalraven-ESG", "CovidAnalytics-DELPHI", "USC-SI_kJalpha", 
            "COVIDhub-4_week_ensemble", "CU-nochange", "Karlen-pypm", "CU-scenario_low", 
            "COVIDhub-ensemble")
if (!file.exists(here("data", "pred_data_20220608.rds"))){
    pred_case <- load_forecasts(
        models = models, 
        dates = d_range, 
        date_window_size = 6,
        locations = state.name, 
        types = "point", 
        targets = inc_case_targets, 
        source = "zoltar", 
        verbose = FALSE, 
        as_of = curr_date, 
        hub = c("US")
    )
    saveRDS(pred_case, file = here("data", "pred_data_20220608.rds"))
} else {
    pred_case <- readRDS(file = here("data", "pred_data_20220608.rds"))
}

ggplot(pred_case, aes(x = model, y = target_end_date)) + geom_point() + 
    coord_flip()
```

Let's load the corresponding truth data  

```{r, cache=TRUE}
truth_data <- load_truth(
    truth_source = "JHU", 
    target_variable = "inc case", 
    locations = state.name
)

true_range <- pred_case %>% pull(target_end_date) %>% 
    unique() %>% 
    lubridate::as_date()

# convert truth_data to epi_df
truth_epidf <- truth_data %>% 
    select(-c(model, location, target_variable, location_name, 
              abbreviation, full_location_name)) %>% 
    dplyr::rename("time_value" = "target_end_date") %>%
    filter(time_value %in% true_range) %>%
    as_epi_df(geo_type = "state")

truth_epidf
```

# Surge classification using relative change growth rate formulation

We define surge for a given date using relative change growth rate formulation times the bandwidth using the implementation from `epiprocess`: 

$$\frac{1}{h} * \left(\frac{\bar{B}}{\bar{A}} - 1\right) = \frac{1}{h} * \left(\frac{\bar{B} - \bar{A}}{\bar{A}}\right) = \\ \frac{1}{h} * \left(\frac{(h)^{-1}\left(\sum_{t = T+1}^{T + h} Y_t - \sum_{t = T+1-h}^{T} Y_t\right)}{(h)^{-1}\sum_{t = T+1-h}^{T} Y_t}\right) = \frac{1}{h} R^{h}_{T + h}$$

A surge is defined for time-point $T$ as the difference in cumulative incident cases between the periods of $T+1$ and $T+h$ and $T$ and $T-h$. As such, an $h$-week ahead forecaster is a nowcaster of whether or not we're currently in a surge. Here, we classify for each week whether it is a surge based on it's $h$-week-ahead and $h$-week-prior data. An upswing is defined as the point at the beginning of a surge. Here we used the threshold of 50\% increase in cumulative difference in the $h$-week-ahead compared to $h$-week-prior data.

```{r}
truth_epidf <- truth_epidf %>% 
    mutate(gr = growth_rate(y = value, method = "rel_change", h = h) * h) %>%
    mutate(surge = case_when(
        gr >= surge_thresh & value >= min_inc ~ TRUE,
        TRUE ~ FALSE
    )) 

truth_epidf <- truth_epidf %>% 
    group_by(geo_value) %>% epi_slide(~{
        bef <- .x$surge[1]
        focal <- .x$surge[2]
        if (is.na(focal)){
            out <- NA
        } else {
            if (bef == FALSE & focal == TRUE){
                out <- TRUE
            } else if (bef == TRUE & focal == FALSE){
                out <- FALSE
            } else {
                out <- FALSE
            }
        }
        return(out)
}, n = 2 * 7 * 1, align = "right", new_col_name = "upswing")

ggplot(truth_epidf %>% filter(geo_value == "al"), aes(x = time_value, y = value)) +
    geom_point(aes(col = surge, shape = upswing), size = 2.5) + geom_line(alpha = 0.5) + 
    labs(x = "Weekly data", y = "Incident cases by week")
```

# Nowcasting surges    

Here we use the following procedures for a focal timepoint $T$ and bandwidth $h$ (for example, 4)  

1. We take a time period from $T+1- h$ to $T+h$. For example at 2021-01-23, we'd be taking the period from 2021-01-02 to 2021-02-20.     
2. We take the truth period to be from $T+1-h$ to $T$ (inclusive) and the forecasting period to be from $T+1$ to $T+h$. The truth period would have real underlying incident cases while the forecasting period has forecast incident cases at times 1-$h$ weeks ahead. Due to the forecasting date being on Monday instead of exactly one week before the proposed target date, we take forecast incident values from the forecast date closest to the time period defined at $T+1$. For example, at 2021-01-23, we would take forecast values for 2021-01-30 onwards from a forecast date of 2021-01-25.  
3. We compute the growth rate at time $T$ using these two periods as per the formula above   
4. We then classify periods as surges using the definition and thresholds defined above.  


```{r}
forecast <- pred_case %>% dplyr::rename("pred" = value, "time_value" = target_end_date)
actual <- truth_epidf %>% dplyr::rename("obs" = value)

combined <- left_join(actual, forecast) %>% 
    select(geo_value, population, geo_type, obs, pred, 
           time_value, forecast_date, model, horizon, surge, upswing)

combined <- combined %>% ungroup() %>% as_epi_df()
head(combined)
```


Let's define the slide function  


```{r}
# this function combines real case counts from time points t-h to t and forecasted 
# case counts from t+1 to t+h. Growth rate at time t is then estimated using the relative change 
# method
mismatch_slide <- function(slide_df, h){
    query_dates <- slide_df %>% pull(time_value) %>% unique() 
    req_len <- h * 2
    # if not enough weeks for prediction
    if (length(query_dates) != req_len){ 
        # this is an exception for when the dates are truncated at the end and beginning of the interval
        out <- NA_real_
    } else {
        # true dates, pred dates and ref dates 
        t_date <- query_dates[1:(req_len - h)]
        ref <- tail(t_date, n = 1)
        p_date <- query_dates[(req_len - h + 1):req_len]
        #print(query_dates)
        #print(ref)
        f_date <- slide_df %>% 
            filter(time_value == head(p_date, n = 1) & horizon == 1) %>%
            pull(forecast_date) %>% unique()
        
        if (length(f_date) == 0){
            # exceptions where there are no relevant forecasts for when the time step 1 ahead of 
            # the reference date
            out <- NA_real_
        } else {
            # get predicted values 
            pred <- slide_df %>% filter(forecast_date == f_date) %>% 
                dplyr::pull(pred)
            
            if (length(pred) < h){
                # this exception is when the model does not forecast $h$ weeks in advance 
                # (e.g. h = 4 but horizon only extends to 3)
                out <- NA_real_
            } else {
                # get true values 
                obs <- slide_df %>% filter(time_value %in% t_date) %>% 
                    select(time_value, obs) %>% 
                    distinct() %>% pull(obs)
                
                # put everything in an epi_df for posterity
                new_df <- tibble(
                    time_value = c(t_date, p_date), 
                    value = c(obs, pred), 
                    geo_value = "placeholder"
                ) %>% as_epi_df()
            
                out <- new_df %>% 
                    mutate(gr_pred = growth_rate(y = value, h = h, method = "rel_change") * h) %>% 
                    filter(time_value == ref) %>% pull(gr_pred)
            }
        }
    }
    return(out)
}
```

Let's use this with `epi_slide` function and loop through all the models  

```{r, cache = TRUE}
begin <- Sys.time()
gradient_calc <- combined %>% 
    group_by(geo_value, model) %>% 
    epi_slide(~mismatch_slide(.x, h = h), n = 2 * 7 * h, align = "center", new_col_name = "pred_gr") %>% 
    dplyr::ungroup() %>%
    dplyr::select(-c(forecast_date, horizon, pred)) %>% distinct()
end <- Sys.time()
print(end - begin)
saveRDS(gradient_calc, file = here("output", "calc_gr.rds"))
```

Classify each time point as either a surge/upswing or not based on established criteria.  

```{r}
gradient_classif <- gradient_calc %>%
    mutate(surge_pred = case_when(
        pred_gr >= surge_thresh & obs >= min_inc ~ TRUE,
        is.na(pred_gr) ~ NA, 
        TRUE ~ FALSE
    )) %>% group_by(geo_value, model) %>% 
    epi_slide(~{
        bef <- .x$surge_pred[1]
        focal <- .x$surge_pred[2]
        if (is.na(focal) | is.na(bef)){
            out <- NA
        } else {
            if (bef == FALSE & focal == TRUE){
                out <- TRUE
            } else if (bef == TRUE & focal == FALSE){
                out <- FALSE
            } else {
                out <- FALSE
            }
        }
        return(out)
    }, n = 2 * 7 * 1, align = "right", new_col_name = "upswing_pred")
    

head(gradient_classif)
```

Perform evaluation  

```{r, message=FALSE}
surge_eval <- map_dfr(models, ~{
    misclass_surge <- gradient_classif %>% ungroup() %>% 
        filter(model == .x) %>% 
        filter(!is.na(surge_pred)) %>% 
        summarise(misclass = mean(surge != surge_pred), 
                  sens = sum(surge & surge_pred)/sum(surge), 
                  spec = sum(!surge & !surge_pred)/sum(!surge))
    pmod <- pROC::roc(surge ~ pred_gr, data = gradient_classif %>% filter(model == .x), 
                      subset = !is.na(surge_pred))
    misclass_surge <- mutate(misclass_surge, auc = round(as.numeric(pmod$auc),3), type = "surge")
    misclass_surge <- misclass_surge %>% mutate(model = .x)
        
})

upswing_eval <- map_dfr(models, ~{
    misclass_upswing <- gradient_classif %>% ungroup() %>% 
        filter(model == .x) %>% 
        filter(!is.na(upswing_pred)) %>% 
        summarise(misclass = mean(upswing != upswing_pred),
                  upswing_prev = mean(upswing),
                  sens = sum(upswing & upswing_pred)/sum(upswing), 
                  spec = sum(!upswing & !upswing_pred)/sum(!upswing))
    pmod <- pROC::roc(upswing ~ pred_gr, data = gradient_classif %>% filter(model == .x), 
                      subset = !is.na(upswing_pred))
    misclass_upswing <- mutate(misclass_upswing, auc = round(as.numeric(pmod$auc),3), type = "upswing")
    misclass_upswing <- misclass_upswing %>% mutate(model = .x)
        
})

saveRDS(upswing_eval, file = here("output", "upswing_eval.rds"))
saveRDS(surge_eval, file = here("output", "surge_eval.rds"))
knitr::kable(surge_eval)
knitr::kable(upswing_eval)
```

Visualize performance values  

```{r}
ggplot(surge_eval, aes(y = misclass, x = reorder(model, -misclass))) + 
    coord_flip() + geom_point() + 
    labs(x = "Models", y = "Misclassification", title = str_wrap("Classifying whether a week is 'surging' based on growth rate estimation", width = 50))
```
```{r}
ggplot(upswing_eval, aes(y = misclass, x = reorder(model, -misclass))) + 
    coord_flip() + geom_point() + 
    labs(x = "Models", y = "Misclassification", title = str_wrap("Classifying whether a week is an upswing or the beginning of the surge period", width = 50))
```

```{r}
library(pals)
colors <- pals::glasbey(n = 9)

suppressMessages(mods <- map(models, ~{
    pROC::roc(upswing ~ pred_gr, data = gradient_classif %>% filter(model == .x), 
                  subset = !is.na(upswing_pred), verbose = FALSE)
}))
names(mods) <- models
plot(mods[[1]], col = colors[1], legacy.axes = TRUE)
for (i in 2:length(mods)){
    plot(mods[[i]], add = TRUE, col = colors[i])
}

legend(0.4,0.5,legend=models, col = colors, lty = 1, cex = 0.8)
```


# Ensemble only evaluation    

```{r}
pred_case <- load_forecasts(
    models = "COVIDhub-ensemble", 
    dates = d_range, 
    date_window_size = 6,
    locations = state.name, 
    types = "point", 
    targets = inc_case_targets, 
    source = "zoltar", 
    verbose = FALSE, 
    as_of = curr_date, 
    hub = c("US")
) 

df <- forecast_surge_eval(source = .x)
# Misclassification rate    
misclass_surge <- df %>% ungroup(geo_value) %>% 
    filter(!is.na(surge_pred)) %>% 
    summarise(misclass = mean(surge != surge_pred), 
              class_prop = mean(surge),
              sens = sum(surge & surge_pred)/sum(surge), 
              spec = sum(!surge & !surge_pred)/sum(!surge))

pmod <- pROC::roc(surge ~ pred_gr, data = df, subset = !is.na(surge_pred))
misclass_surge <- mutate(misclass_surge, auc = round(as.numeric(pmod$auc),3), type = "surge")

misclass_upswing<- df %>% ungroup(geo_value) %>% 
    filter(!is.na(upswing_pred)) %>% 
    summarise(misclass = mean(upswing != upswing_pred), 
              class_prop = mean(upswing),
              sens = sum(upswing & upswing_pred)/sum(upswing), 
              spec = sum(!upswing & !upswing_pred)/sum(!upswing))

pmod <- pROC::roc(upswing ~ pred_gr, data = df, subset = !is.na(upswing_pred))
misclass_upswing <- mutate(misclass_upswing, auc = round(as.numeric(pmod$auc),3), type = "upswing")
results <- bind_rows(misclass_upswing, misclass_surge)
print(results)
```

Plotting incident cases and surge classification   

```{r}
p1 <- ggplot(df, aes(x = time_value, y = value)) +
    geom_point(aes(col = surge)) + 
    geom_line(alpha = 0.5) + 
    labs(x = "Week", col = "Surge classification from real data", y = "Incident cases")
p2 <- ggplot(df, aes(x = time_value, y = value)) + 
    geom_point(aes(col = surge_pred)) + 
    geom_line(alpha = 0.5) + 
    labs(x = "Week", col = "Nowcasting surges", y = "Incident cases")

p1 /p2

```


```{r}
p1 <- ggplot(df, aes(x = time_value, y = value)) +
    geom_point(aes(col = surge)) + 
    geom_line(alpha = 0.5) + 
    labs(x = "Week", col = "Surge classification from real data", y = "Incident cases")
p2 <- ggplot(df, aes(x = time_value, y = value)) + 
    geom_point(aes(col = surge_pred)) + 
    geom_line(alpha = 0.5) + 
    labs(x = "Week", col = "Nowcasting surges", y = "Incident cases")

p1 /p2

```


