---
title: "Comparing surge definition to CDC"
author: "Quang Nguyen"
date: "Last compiled on `r format(Sys.time(), '%Y-%m-%d')`"
output: 
    #github_document: default
    html_document:
        keep_md: yes
        toc: true
        toc_float: true
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data pre-loading and processing

```{r, message=FALSE, warning=FALSE}
library(here)
library(covidcast)
library(epiprocess)
library(tidyverse)
library(ggsci)
library(tsibble)
library(patchwork)
here::i_am(path = "notebooks/upswing-CDC.Rmd")
source(here("R", "utils.R"))
# (settings <- get_settings(start_date = "2020-06-01", end_date = "2022-03-01"))
```

Loading CDC Rt estimates

```{r cdc_load}
cdc_rt <- read_csv(file = here("data", "cdc_rt.csv")) %>% select(-1)
cdc_gr <- read_csv(file = here("data", "cdc_gr.csv")) %>% select(-1)

start_date <- as.character(min(cdc_rt$date))
end_date <- as.character(max(cdc_rt$date))
h <- 4
surge_thresh <- 0.5
min_inc <- 20


head(cdc_rt)
head(cdc_gr)

```

Obtain case-count data from JHU using start and end dates as defined by the CDC file which is from `r start_date` and `r end_date`.

```{r}

print(paste("Computing surge data from", start_date, "to", end_date))
# load data from covid_cast signal
df <- covidcast_signal(data_source = "jhu-csse", 
                          signal = "confirmed_incidence_num", 
                          start_day = start_date, 
                          end_day = end_date, 
                          geo_type = "state",
                          as_of = Sys.Date()) 
```

Here, we convert `covidcast_signal` data into `epi_df` and `tsibble`. We have to back-convert to `epi_df` to work with `growth_rate` function.

```{r}
# convert to epi-df 
# after temporal operations in tsibble have to back-convert to epi_df
df <- df %>% 
        as_epi_df(geo_type = "state", time_type = "day", as_of = max(.$issue)) %>% 
        select(geo_value, time_value, cases = value) %>% as_tsibble() %>%
        index_by(epiweek = ~yearweek(., week_start = 7)) %>% 
        group_by(geo_value) %>% 
        summarise(cases = sum(cases)) %>% ungroup() %>% 
        as_tibble() %>% dplyr::rename(time_value = epiweek) %>% 
        as_epi_df(geo_type = "state", time_type = "week") %>%
        mutate(geo_value = covidcast::abbr_to_name(geo_value, ignore.case = TRUE))

head(df)
```

# Surge/Downswing computation

Compute upswing using growth_rate from epiprocess and use approximate labels. For surge classification using growth rate, we use a the weekly percent increase threshold in incidence to be 50% or more assuming that the incident case per week is at least 20. For surge classification using Rt, classified a week as increasing if the Rt had a 95% probability of being greater than or less than 1 (weekly Rt estimates were obtained by averaging daily estimates across the entire week).

```{r}
df <- df %>% group_by(geo_value) %>% 
    mutate(gr = growth_rate(y = cases, method = "rel_change", h = h) * h) %>% 
    mutate(surge = case_when(
        gr >= surge_thresh & cases >= min_inc ~ "increasing",
        gr <= -surge_thresh & cases >= min_inc ~ "decreasing", 
        TRUE ~ "unclassified"
    ))

rt <- cdc_rt %>% 
    mutate(time_value = yearweek(date, week_start = 7)) %>% 
    rename("geo_value" = "state") %>% 
    group_by(geo_value, time_value) %>% 
    summarise(rt = mean(mean), rt_upper = mean(upper_90), rt_lower = mean(lower_90)) %>% 
    mutate(surge = case_when(
        rt_lower > 1 ~ "increasing", 
        rt_upper < 1 ~ "decreasing",
        TRUE ~ "unclassified"
    ))

head(df)

head(rt)


joint_df <- inner_join(df, rt, by=c("geo_value", "time_value")) %>% 
    pivot_longer(c(surge.x, surge.y), names_to = "source", values_to = "surge") %>%
    mutate(source = if_else(source == "surge.x", "gr", "cdc"))

cols <- c("unclassified" = "#003f5c", "increasing" = "#bc5090", "decreasing" = "#ffa600")
```

# Per state plots of increasing/decreasing/unclassified {.tabset}

These plots represent per-week classification based on the schema above. As such, some weeks are considered increasing/decreasing while others are "unclassified"

```{r, results='asis'}

for (i in joint_df %>% pull(geo_value) %>% unique()){
    cat('##', i, '\n')
    p1 <- ggplot(joint_df %>% filter(geo_value == i) %>% filter(source == "gr"), 
                 aes(x = time_value, y = cases)) + 
        geom_line(alpha = 0.5) + 
        geom_point(aes(col = surge)) + 
        geom_rect(aes(ymin = 0, ymax = max(cases) + 10, xmin = time_value - 0.5, 
                      xmax = time_value + 0.5, fill = surge), alpha = 0.3) + 
        scale_color_manual(values = cols) + 
        scale_fill_manual(values = cols) + 
        theme_bw() +
        labs(y = "Cases", x = "Epiweek (Year-Week)", col = "Phase classification", fill = "Phase classification", 
             title = str_wrap("Phase classification via growth rate thresholded at 0.5", width = 30))
    p2 <- ggplot(joint_df %>% filter(geo_value == i) %>% filter(source == "cdc"), 
                 aes(x = time_value, y = rt)) +
        geom_point() + 
        geom_line(alpha = 0.5) + 
        geom_point(aes(col = surge)) + 
        geom_ribbon(aes(ymin = rt_lower, ymax = rt_upper), alpha = 0.4) +
        geom_rect(aes(ymin = 0.5, ymax = max(rt) + 0.3, xmin = time_value - 0.5, 
                      xmax = time_value + 0.5, fill = surge), alpha = 0.3) + 
        scale_color_manual(values = cols) + 
        scale_fill_manual(values = cols) + 
        theme_bw() +
        labs(y = "Rt", x = "Epiweek (Year-Week)", col = "Phase classification", fill = "Phase classification", 
             title = str_wrap("Phase classification via 95% CI estimates of Rt", width = 30))

    
    x <- joint_df$gr[joint_df$geo_value == i]
    y <- joint_df$rt[joint_df$geo_value == i]
    
    y <- y[-which(is.na(x))]
    x <- na.omit(x)
    
    p3 <- ggplot(joint_df %>% filter(geo_value == i), 
                 aes(x = rank(gr), y = rank(rt))) + 
        geom_point() + 
        labs(x = "rank(growth_rate)", y = "rank(Rt)", 
             title = paste("Spearman correlation:", round(cor(x,y, method = "spearman"), 3))) +
        theme_bw()
    
    layout <- "
    AAA#
    AAA#
    AAAC
    BBBC
    BBB#
    BBB#
    "
        
    print(p1 + p2 + p3 + plot_layout(guides = "collect", design = layout))
    cat("\n \n")
}

```

# Per-state plots with full phase classification 

These plots represent per-state classifications with include full phase classifications. Specifically:

- For weeks between two increasing phases, we classified them as increasing and for weeks between two decreasing periods, we classified them as decreasing.  
- Weeks between increasing and decreasing phases were classified as peaks, whereas nadirs were defined as periods between decreasing and increasing phases.  
- If peaks or nadirs only contained an interval of 1 week, we included the week before and after in the peak/nadir classification.
- Periods at the beginning or the end of the analysis period were classified as a continuation of whichever phase preceded or followed them.    


```{r}
# phase classification was done by counting all "unclassified" as missing 
# we then use count_gaps in order to define all complete phases and then assign labels to them  
# and then loop through the main data frame and assign labels based on gaps (if surge is unclassified)
# for some reason using map functions doesn't work when trying to filter time_value >= .from and <= .to

get_lab <- function(df, geo, time){
    return(df %>% filter(geo_value == geo) %>% 
               filter(time_value == time) %>%
               pull(surge))
}

phase_clasification <- function(src){
    data <- joint_df %>% filter(source == src)
    gaps <- data %>% filter(surge != "unclassified") %>% 
        as_tsibble(key = geo_value) %>% tsibble::count_gaps(.full = TRUE) 
    
    
    gaps <- gaps %>% mutate(label = pmap_chr(gaps, function(geo_value, .from, .to, ...){
        before <- get_lab(data, geo_value, .from - 1)
        after <- get_lab(data, geo_value, .to + 1)
        if (length(before) == 0){
            val <- after
        } else if (length(after) == 0){
            val <- before
        } else {
            val <- case_when(
                before == "increasing" & after == "decreasing" ~ "peaks",
                before == "decreasing" & after == "increasing" ~ "nadirs",
                before == "increasing" & after == "increasing" ~ "increasing",
                before == "decreasing" & after == "decreasing" ~ "decreasing",
                TRUE ~ NA_character_
            )
        }
        return(val)
    }))
    
    phases <- vector(length = nrow(data))
    for (i in seq_along(phases)){
        if (data[i,]$surge != "unclassified"){
            lab <- data[i,]$surge
        } else {
             t <- data[i,]$time_value
             g <- data[i,]$geo_value
             lab <- gaps %>% filter(geo_value == g) %>% 
                 filter(.from <= t) %>% 
                 filter(.to >= t) %>% pull(label) %>% as.vector()
             if (length(lab) == 0){
                 lab <- NA_character_
             }
        }
        phases[i] <- lab
    }
    data$phases <- phases
    return(data)
}
```

Applying phase classification to CDC Rt and case count data sets  

```{r}
gr_phase <- phase_clasification("gr")
rt_phase <- phase_clasification("cdc")
cols <- c("peaks" = "#003f5c", "increasing" = "#bc5090", "decreasing" = "#ffa600", "nadirs" = "#58508d")
```

## Per-state plots {.tabset}

```{r, results='asis'}
for (i in joint_df %>% pull(geo_value) %>% unique()){
    cat('###', i, '\n')
    p1 <- ggplot(gr_phase %>% filter(geo_value == i), 
                 aes(x = time_value, y = cases)) + 
        geom_line(alpha = 0.5) + 
        geom_point(aes(col = phases)) + 
        geom_rect(aes(ymin = 0, ymax = max(cases) + 10, xmin = time_value - 0.5, 
                      xmax = time_value + 0.5, fill = phases), alpha = 0.3) + 
        scale_color_manual(values = cols) + 
        scale_fill_manual(values = cols) + 
        theme_bw() +
        labs(y = "Cases", x = "Epiweek (Year-Week)", col = "Phase classification", fill = "Phase classification", 
             title = str_wrap("Phase classification via growth rate thresholded at 0.5", width = 30))
    p2 <- ggplot(rt_phase %>% filter(geo_value == i), 
                 aes(x = time_value, y = rt)) +
        geom_point() + 
        geom_line(alpha = 0.5) + 
        geom_point(aes(col = phases)) + 
        geom_ribbon(aes(ymin = rt_lower, ymax = rt_upper), alpha = 0.4) +
        geom_rect(aes(ymin = 0.5, ymax = max(rt) + 0.3, xmin = time_value - 0.5, 
                      xmax = time_value + 0.5, fill = phases), alpha = 0.3) + 
        scale_color_manual(values = cols) + 
        scale_fill_manual(values = cols) + 
        theme_bw() +
        labs(y = "Rt", x = "Epiweek (Year-Week)", col = "Phase classification", fill = "Phase classification", 
             title = str_wrap("Phase classification via 95% CI estimates of Rt", width = 30))
    print(p1 / p2 + plot_layout(guides = "collect"))
    cat("\n\n")
}
```





